### 2015-08-17 submission20150818/plos_template.tex
add submission20150818

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'').
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechansim, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Figs. S2C S2D).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figs. 6B, 6C ,S6C and S6D).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'', \cite{pmid23318258}). 
Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S3A and S3B), and signature 2 disappeared when we removed this sample (see Fig. S4).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Figs. S5A, S5B and S5C).
However, for each cancer type, only a few samples support this signature (see Figs. S5D, S5E and S5F),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. The 12 inferred signatures are highly consistent among cancer types except for B-cell lymphoma (see Fig. S2A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S2A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

Signatures 10, 11, 12, 19 and 21 provide further examples
of our method refining previously-observed signatures, highlighting strand biases and/or sequence context effects,
particularly 2 bases upstream of the substitution.
Signature 10 (C $>$T at [CT]pCpC) was observed in head and neck cancers and melanomas, and probably relates to ultraviolet light.
Consistent strand specificities among the two cancer types (Fig. S2E) matches previous results \cite{pmid23318258}, 
but our analysis additionally highlights elevated abundance of T at the $-2$ position (Figs. 6D and S2E). 
Signature 11 (C $>$ T at GpCp[CG]) appears in small-cell lung cancers and stomach cancers,
and seems to be the same as ``signature 15'' in the previous study, whose function remains unclear. Again our analysis highlights elevated abundance of G at the $-2$ position (Figs. 6E and S2F).
Signatures 12 (C $>$ T at [CG]pCp[CT]), 19 (T $>$ C at GpTpN) and 21 (T $>$ [CG] at CpTpT) observed in pilocytic astrocytomas, stomach cancers and oesophagus cancers, respectively, agree well with those detected in the same cancer types in the previous study \cite{pmid23318258}. However our analysis again refines these signatures,
highlighting a strand bias in all three, and sequence context effects at the -2 position in Signatures 12 and 21.
%although the causes of these signatures remain unknown.
%which should be validated by analysis of more samples from different cohorts and cancer types.

@@

The remaining signatures largely recapitulate previous results.
Signature 3 and 5 (C $>$ A at NpCpN) observed in head-and-neck cancers and three types of lung cancers are probably associated with tobacco smoking.
The estimated signature in each cancer type shows higher mutation prevalence on the template strand (Fig. S2B), 
which is consistent with the previous study \cite{pmid12379884, pmid23318258}.
Signature 6 (C $>$ A at NpCp[AT]) observed in neuroblastomas matches the pattern detected in the same cancer type in the previous study.
Signature 7 (C $>$ T at NpCpG sites) was observed in 25 out of 30 cancer type, and arguably relates to deamination of 5-methyl-cytosine.
Signature 9 (C $>$ T at NpCp[CT]) was observed in melanomas and glioblastomas, and is probably associated with a chemotherapy drug, temozolomide.
Signature 18 (T $>$ C at ApTp[AG]) observed in liver cancers has been shown to be more common in Asian cases than in other ancestries \cite{pmid25362482}, 
though the source of this signature is still not clear. 
In this signature, we observe a very strong strand specificity as shown in \cite{pmid25362482, pmid23318258},
suggesting a possible role for transcription-coupled nucleotide excision repairs.

@@

\subsection*{S2 Figure}
\label{nature2013_sig_list}
{\bf The list of several signatures extracted in each cancer type in the Alexandrov et al. (2013) data.}
(A) APOBEC signatures (signature 13) obtained in each cancer type.
(B) Smoking signature in each cancer type.
(C) The first Pol $\epsilon$ signature (signature 1)  in each cancer type.
(D) The second Pol $\epsilon$ signature (signature 8) in each cancer type.
(E) The ultraviolet signature i(signature 10) n each cancer type.
(F) Unknown signature (signature 11) obtained in lung small cell carcinomas and stomach cancers.

\subsection*{S3 Figure}
\label{LGG_membership}
{\bf The estimated membership parameters of low grade gliomas.}
(A, B) Estimated membership parameter by the proposed method in normal and log scale.
We have selected top 100 cancer samples according to the number of mutation.
The height of bar shows (the logarithm of) the number of mutations for each sample,
and the ratio of colored division shows the ratio of estimated membership parameters for each signature and sample.
The low grade glioma specific signature detected by the proposed method is the signature 2.
We can see that the mutations corresponding to signature 2 is mostly from the sample with an extremely high mutation rate.

\subsection*{S4 Figure}
\label{LGG_signatures}
{\bf The signatures obtained for the original data and for the data without the hyper-mutated case.}
(A) The result for the original data ($K = 3$). The first (from the left) signature seems to be one from deamination of 5-methyl-cytosine. 
The second signature is the low grade glioma signature.
(B, C, D, E) The result for the data without the hyper-mutated sample for $K = 2, 3, 4$ and $5$, respectively.
Although the signature related to deamination of 5-methyl-cytosine remained, low grade glioma specific signature could not be observed.

\subsection*{S5 Figure}
\label{oxidation}
{\bf The putative oxidative artifact signatures and membership parameters estimated for each cancer.}
(A) The second signature detected in kidney clear cell carcinomas.
(B) The first signature detected in lung adenocarcinomas.
(C) The first signature detected in melanomas
(D, E, F) Estimated membership parameter for kidney clear cell carcinomas, lung adenocarcinomas and melanomas, respectively.
For each cancer type, 
We have selected top 100 cancer samples according to the number of mutation.
The height of bar shows the number of mutations for each sample,
and the ratio of colored division shows the ratio of estimated membership parameters for each signature and sample.
We can see that the signature corresponding to putative oxidative artifacts concentrates on a small number of samples.

>>>>

Figs. \ref{nature2013_sig_summary} and \ref{nature2013_sig_member} summarise the results.
In total, we identified 27 mutation signatures.
Many of these signatures show reassuring similarities with
signatures identified in previous studies. However
signatures from our independence model, because of its ability to effectively
and parsimoniously deal with both $\pm 2$ flanking
base context and strand bias, are often more refined, highlighting additional details or features not previously evident. By comparing the composition of nucleotides and cancer types exhibiting the signatures with results of previous studies, we were able to associate many of the detected signatures with known mutational processes.
In addition, as we reviewed these signatures
and compared them with previous work, we noticed connections
that, while not directly related to our new model, appear
novel and noteworthy. The remainder of this section
provides a comprehensive discussion of these findings.

@@

Signatures 1 and 8 (C $>$ A at TpCpT and C $>$ T at TpCpG, respectively) observed in colorectal and uterine cancers
appear likely to be associated with deregulated activity of the error-prone polymerase Pol $\epsilon$. 
In previous analyses of these data \cite{pmid23945592}, 
the signature for Pol $\epsilon$ dysfunction was represented by a single signature (their ``signature 10'', see Fig. S2).
In contrast our new approach uses two signatures.
Since these signatures are highly correlated,
and appear connected by
a single biological mechansim, we certainly do not argue
that inferring them as a single signature is ``wrong". However, splitting them into two signatures does help highlight certain features. Specifically,
signature 1 shows a transcription strand bias whereas signature 8 does not, and this is true for both colorectal and uterus cancers (Figs. S3C and S3D).
This strand bias may be connected with the enrichment of 
C $>$A at TpCpT mutations in leading strands of replication forks
observed by \cite{pmid25228659}. Although replication strand bias is different from transcription strand bias, these two biases may be connected through the fact that replication origins prefer transcription start sites \cite{pmid23187890}.

These signatures also illustrate the ability of our model
to help highlight sequence context effects beyond the immediate flanking bases. Specifically, both signatures 1 and 8  show an elevated frequency of the T base at position $-2$,
and signature 1 also shows slightly elevated frequency of the T base at position $+2$
 (Figs. 6B, 6C ,S3C and S3D).
A previous study of Pol $\epsilon$  \cite{pmid25228659} 
found that a nonsense mutation R23X of TP53 is enriched in cancers with Pol $\epsilon$ defects.
In fact, the pattern of this mutation is C $>$ T at TpTpCpGpA, closely matching signature 8.
This illustrates that the inclusion of $\pm 2$ bases 
into signatures may be helpful for identifying underlying mechanisms.

@@

%XX this is important that it is drive by a single sample, but maybe just because yuichi looked more carefully and not new method XX
Signature 2 (C $>$ A at [CT]pCpT) is observed solely in low grade gliomas, and
appears related to, but slightly different from, the signature previously detected in the same cancer types (``signature 14'', \cite{pmid23318258}). 
Indeed, the corresponding signature in the previous study shows very complex patterns (C $>$ A at NpCpT or C $>$ T at GpTpN). Further investigation
revealed that this signature is driven by 
a single sample with an extremely high mutation rate (see Figs. S4A and S4B), and signature 2 disappeared when we removed this sample (see Fig. S5).
It may be that the complex low-grade-glioma specific signature detected in the previous study is driven by the same single sample.
We suggest that these signatures should be treated with caution until validated in additional samples. 
% XXX Need to discuss this wording and what it means XXX

%XX Also important, making a connection, but not the new method XX
Signature 4 (C $>$ A at CpCpG) observed in kidney clear cell carcinomas, lung adenocarcinomas and melanomas 
seems to correspond to the ``signature R2''  detected in the same cancer types (plus lung squamous carcinomas) in 
\cite{pmid23318258} (see their Supplementary Figures).
Again our analysis highlights additional contextual information, with a strikingly elevated frequency of base C at the $-2$ position (Figs. S6A, S6B and S6C).
However, for each cancer type, only a few samples support this signature (see Figs. S6D, S6E and S6F),
and the corresponding signature 
could not be validated in the previous study: most somatic mutations corresponding to that signature could not be validated by re-sequencing or visual inspection of BAM files using genomic viewers. Again, further investigation
yielded a potential explanation for this finding:
this signature largely matches that of a putative artifact caused by oxidation of DNA during acoustic shearing \cite{pmid23303777}, 
and we conclude that this signature, and the corresponding signature in previous work, are likely artefactual.
Although not of direct biological interest,
identifying artefactual signatures could be helpful in removing false positive mutations.

%XX highlights robustness of approach XX
Signature 13 (T $>$ [AGT] at TpCpN sites) was observed in 12 cancer types, and is surely related to the activity of the APOBEC family. The 12 inferred signatures are highly consistent among cancer types except for B-cell lymphoma (see Fig. S3A), highlighting the robustness
of our approach.  Almost all of them show enrichment of A and T and depletion of G base at the $-2$ position (Fig. 6A and S3A), 
consistent with the UCUT data above and previous analyses \cite{pmid23318258}.
The estimated transcribed strand specificities varied among cancer types, suggesting that there is not consistent strand-specificity in APOBEC signatures (and the observed variation may be due to estimation errors).
%Yuichi is that true?
Signatures 15 and 16 may also be related to APOBEC, although the estimated signatures are sufficiently different from 13 that they
were not merged into a single 
cluster by our specified clustering criteria.

Signatures 10, 11, 12, 19 and 21 provide further examples
of our method refining previously-observed signatures, highlighting strand biases and/or sequence context effects,
particularly 2 bases upstream of the substitution.
Signature 10 (C $>$T at [CT]pCpC) was observed in head and neck cancers and melanomas, and probably relates to ultraviolet light.
Consistent strand specificities among the two cancer types (Fig. S3E) matches previous results \cite{pmid23318258}, 
but our analysis additionally highlights elevated abundance of T at the $-2$ position (Figs. 6D and S3E). 
Signature 11 (C $>$ T at GpCp[CG]) appears in small-cell lung cancers and stomach cancers,
and seems to be the same as ``signature 15'' in the previous study, whose function remains unclear. Again our analysis highlights elevated abundance of G at the $-2$ position (Figs. 6E and S3F).
Signatures 12 (C $>$ T at [CG]pCp[CT]), 19 (T $>$ C at GpTpN) and 21 (T $>$ [CG] at CpTpT) observed in pilocytic astrocytomas, stomach cancers and oesophagus cancers, respectively, agree well with those detected in the same cancer types in the previous study \cite{pmid23318258}. However our analysis again refines these signatures,
highlighting a strand bias in all three, and sequence context effects at the -2 position in Signatures 12 and 21.
%although the causes of these signatures remain unknown.
%which should be validated by analysis of more samples from different cohorts and cancer types.

@@

The remaining signatures largely recapitulate previous results.
Signature 3 and 5 (C $>$ A at NpCpN) observed in head-and-neck cancers and three types of lung cancers are probably associated with tobacco smoking.
The estimated signature in each cancer type shows higher mutation prevalence on the template strand (Fig. S3B), 
which is consistent with the previous study \cite{pmid12379884, pmid23318258}.
Signature 6 (C $>$ A at NpCp[AT]) observed in neuroblastomas matches the pattern detected in the same cancer type in the previous study.
Signature 7 (C $>$ T at NpCpG sites) was observed in 25 out of 30 cancer type, and arguably relates to deamination of 5-methyl-cytosine.
Signature 9 (C $>$ T at NpCp[CT]) was observed in melanomas and glioblastomas, and is probably associated with a chemotherapy drug, temozolomide.
Signature 18 (T $>$ C at ApTp[AG]) observed in liver cancers has been shown to be more common in Asian cases than in other ancestries \cite{pmid25362482}, 
though the source of this signature is still not clear. 
In this signature, we observe a very strong strand specificity as shown in \cite{pmid25362482, pmid23318258},
suggesting a possible role for transcription-coupled nucleotide excision repairs.

@@

\subsection*{S2 Figure}
\label{nature2013_POLE}
{\bf The pol epsilon signature (Signature 10) derived in the Alexandrov et al., (2013).}
The barplots are divided by 6 substitution pattern. 
In each division, 16 bars show joint probabilities of substitution pattern, 5' and 3' bases (ApNpA, ApNpC, ApNpG, ApNpT, $\cdots$, TpNpT).

\subsection*{S3 Figure}
\label{nature2013_sig_list}
{\bf The list of several signatures extracted in each cancer type in the Alexandrov et al. (2013) data.}
(A) APOBEC signatures (signature 13) obtained in each cancer type.
(B) Smoking signature in each cancer type.
(C) The first Pol $\epsilon$ signature (signature 1)  in each cancer type.
(D) The second Pol $\epsilon$ signature (signature 8) in each cancer type.
(E) The ultraviolet signature i(signature 10) n each cancer type.
(F) Unknown signature (signature 11) obtained in lung small cell carcinomas and stomach cancers.

\subsection*{S4 Figure}
\label{LGG_membership}
{\bf The estimated membership parameters of low grade gliomas.}
(A, B) Estimated membership parameter by the proposed method in normal and log scale.
We have selected top 100 cancer samples according to the number of mutation.
The height of bar shows (the logarithm of) the number of mutations for each sample,
and the ratio of colored division shows the ratio of estimated membership parameters for each signature and sample.
The low grade glioma specific signature detected by the proposed method is the signature 2.
We can see that the mutations corresponding to signature 2 is mostly from the sample with an extremely high mutation rate.

\subsection*{S5 Figure}
\label{LGG_signatures}
{\bf The signatures obtained for the original data and for the data without the hyper-mutated case.}
(A) The result for the original data ($K = 3$). The first (from the left) signature seems to be one from deamination of 5-methyl-cytosine. 
The second signature is the low grade glioma signature.
(B, C, D, E) The result for the data without the hyper-mutated sample for $K = 2, 3, 4$ and $5$, respectively.
Although the signature related to deamination of 5-methyl-cytosine remained, low grade glioma specific signature could not be observed.

\subsection*{S6 Figure}
\label{oxidation}
{\bf The putative oxidative artifact signatures and membership parameters estimated for each cancer.}
(A) The second signature detected in kidney clear cell carcinomas.
(B) The first signature detected in lung adenocarcinomas.
(C) The first signature detected in melanomas
(D, E, F) Estimated membership parameter for kidney clear cell carcinomas, lung adenocarcinomas and melanomas, respectively.
For each cancer type, 
We have selected top 100 cancer samples according to the number of mutation.
The height of bar shows the number of mutations for each sample,
and the ratio of colored division shows the ratio of estimated membership parameters for each signature and sample.
We can see that the signature corresponding to putative oxidative artifacts concentrates on a small number of samples.
